A performance assessment of evolutionary algorithms in a volunteer computing system

### 2016-03-13 PPSN2016/paper.Rnw
First full version
Refs #109 #108 #107

Please all revise the paper. I would like to submit it next week to go on to another paper for PPSN. @fergunet @unintendedbear @mariosky revise English, conclusions, everything...

In this paper our focus will be in the second aspect: we will try to
design a decentralized system that, at the same time, is able to use
all resources available. This design will be done incrementally by
changing the system on the client and the server and measuring its
impact on the overall performance: time and evaluations needed to find
the solution. Eventually, we want to find a system that, whatever the
number of users available to perform the experiment, is able to
maximize their contribution to the evolutionary algorithm, at the same
time that the evolutionary algorithm itself makes the most of those
contributions, independently of how many they are.

@@

graph2 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 0.0520509 & 36 & 73 & 129.04545\\
\hline
reboot & 796214.0 & 9 & 6 & 646.0 & 67.0 & 0.0955121 & 45 & 168 & 75.88889\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 0.1575092 & 55 & 225 & 43.28571\\
\hline
\end{tabular}
%% \end{tabular}
%% Initially generated with this code, 
% <<experiments, cache=FALSE,echo=FALSE>>=
%experiment.aggregates$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%kable(experiment.aggregates)
%@ %def 
\end{center}
\end{table}
%
%
\begin{table}
  \caption{Experiments with overlaps eliminated:  summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=3>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Ranked and normalized number of contributions (PUTs) per IP,
  plotted in a log $y$ scale.\label{fig:zipf}}
\label{fig:zipf}
\end{figure}

Since the point of using volunteer computing system is to cut costs,
all experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to \$0.00. In fact, the NodIO system can be deployed to any
system as long as it can run node.js and access to the filesystem, for
log files, is available. All these logs are available under
rrequest. Processed files are available in the same repository that
hosts this paper, as well as the scripts needed to process them.

Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP order) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that is slightly superior to the rest, specially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

@@

So we will have to check whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which reasonably the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimun as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that use less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive  effect of
increasing the contribution of clients to the common pool. However, we
should remark that still the clients will stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact in the algorithm.

@@

% TODO

@@

This work has been supported in part by TIN2014-56494-C4-3-P (Spanish Ministry of Economy and Competitivity),
PROY-PP2015-06 (Plan Propio 2015 UGR)  % please, remember to include all the current projects   ;)   Thanks!
. We would also like to thank the
anonymous reviewers of previous versions of this paper who have really
helped us to improve 
this paper (and our work) with their suggestions. We are also grateful
to Anna S\'aez de Tejada for her help with the data processing
scripts. We are also grateful to {\tt @otisdriftwood} for his help
gathering users for the new experiments.

>>>>

In this paper our focus will be in the second aspect: we will try to
design a decentralized system that, at the same time, is able to use
all resources available. This design will be done incrementally by
changing the system on the client and the server and measuring its
impact on the overall performance: time and evaluations needed to find
the solution. Eventually, we want to find a system that, whatever the
number of users available to perform the experiment, is able to
maximize their contribution to the evolutionary algorithm, at the same
time that the evolutionary algorithm itself makes the most of those
contributions, independently of how many they are, and is able to find
the solution to the problem in less time and with less contributions.

@@

graph2 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 0.0520509 & 36 & 73 & 129.04545\\
\hline
reboot & 796214.0 & 9 & 6 & 646.0 & 67.0 & 0.0955121 & 45 & 168 & 75.88889\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 0.1575092 & 55 & 225 & 43.28571\\
\hline
\end{tabular}
%% \end{tabular}
%% Initially generated with this code, 
% <<experiments, cache=FALSE,echo=FALSE>>=
%experiment.aggregates$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%kable(experiment.aggregates)
%@ %def 
\end{center}
\end{table}
%
%
%
Since the point of using volunteer computing system is to cut costs,
all experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to \$0.00. In fact, the NodIO system can be deployed to any
system as long as it can run node.js and access to the filesystem, for
log files, is available. All these logs are available under
rrequest. Processed files are available in the same repository that
hosts this paper, as well as the scripts needed to process them.

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=3>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Ranked and normalized number of contributions (PUTs) per IP,
  plotted in a log $y$ scale.\label{fig:zipf}}
\label{fig:zipf}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP order) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that is slightly superior to the rest, specially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

@@

%
\begin{table}
  \caption{Experiments with overlaps eliminated:  summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
So we will have to check whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which reasonably the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimun as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that use less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive  effect of
increasing the contribution of clients to the common pool. However, we
should remark that still the clients will stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact in the algorithm.

@@

Our main intention in this paper was to focus on the algorithm itself
so that it became as effective as possible, and able to use
computation by volunteers in the best possible way, or at least better
than the initial configuration. This initial configuration included
{\em overlaps} or clients running islands that had actually started in
a previous experiment in the set. If we are interested in measuring
how many users or their behavior in bulk this is not so important, but
in this case our focus was on the algorithm so we had to try that all
clients participating in it started more or less at the same time and
from a random population. In fact, we managed to do so, except in some
byzantine cases, by making the clients restart when the cache size
collapses to 1 or to a size much smaller than the current; we called
this client configuration {\tt reboot} and {\tt cachecrash}.  This
configuration achiever a positive secondary effect: all clients
contribute to fill the cache, having thus an actual effect on the
chromosome pool and effectively making all clients present contribute
to the simulation.

The last configuration managed to find, with a relatively small number
of evaluations, solutions in half the time that in previous sets of
experiments. Although in the {\tt base} and {\tt graph} the median
time and number of contributions was smaller, we cannot disregard the
possibility that many of those solutions were found thanks to
contributions from {\em carried over} islands with an algorithm that
has run a big amount of evaluations already. At any rate, the {\tt
  cachereboot} finds on average solutions four times as fast as {\tt
  graph2} with twice as many average volunteers (or three times as
many if we consider {\em effective} volunteers), needing 50\% less
evaluations, thus meaning a big improvement in the performance of the
evolutionary algorithm, which was the target of this paper. This
improvement is less with respect to {\tt reboot}, but in any case the
method to detect the beginning of a new run is much more effective and
will be kept in the future.

This future holds many challenges which once again will go in two
different directions: gathering more users and improving the
efficiency of the algorithm run in an asynchronous way by the
users. In the first case, it will be interesting to make it more {\em
  social} by comparing local performance to other's contributions, and
showing it in a graph or being able to tweet advances or the fact that
the solution has been found. Anything that keeps the users running the
system and attracts new users will contribute to the speed of finding
the solution. And the algorithm will have to be improved, mainly by
making the population more diverse. Right now there is a rigid policy
of contribution to the pool: send the best individual, but this could
be relaxed. Some other policies could also be tested, as well as
running heterogenous algorithms in the same way that was done in
\cite{DBLP:journals/grid/ValdezTGVO15}. Diversity, as in any other
evolutionary algorith, seems to be the key, but there are many ways of
approaching it.

@@

This work has been supported in part by TIN2014-56494-C4-3-P (Spanish
Ministry of Economy and Competitivity), PROY-PP2015-06 (Plan Propio
2015 UGR) . We would also like to thank the anonymous reviewers of
previous versions of this paper who have really helped us to improve
this paper (and our work) with their suggestions. We are also grateful
to Anna S\'aez de Tejada for her help with the data processing
scripts. We are also grateful to {\tt @otisdriftwood} for his help
gathering users for the new experiments.
